---
title: "DEM 7283 - Example 1 - Introduction to R"
author: "coreysparks"
date: "January 12, 2014"
output: html_document
---

Welcome to R. R is very different from SAS. R is an interpreted language, not a compiled one. This means, you type something into R and it does it. There is no data step. There are no procs. The SAS and R book is very useful for going between the two programs.  R uses libraries to do different types of analysis, so we will need to install lots of different libraries to do different things. These need to be downloaded from the internet, using the `install.packages()` command. You only need to install a package once. E.g.

`install.packages("lme4")`
will install the lme4 library. To use the functions within it, type

`library(lme4)`

Now you have access to those fuctions. 

Below we will go through a simple R session where we basically review much of what we did last semester. We will load a dataset, print some cases, do some descriptice statistics and plots, t tests, and some linear models and diagnostics of those models.

```{r}
#Load some libraries that I need
library(lmtest)
library (car)
library(Hmisc)
library(sandwich)
library(multcomp)
library(knitr)

#Read in a Commma separated values file, In this case, I am using the Population Reference Bureau's Population Data sheet from 2008.
dat<-read.csv("~/Google Drive/dem7283/PRB2008_All.csv")


#print all of the variable names in the data set
names(dat)


#look at the first 5 cases
head(dat, n=5)

#Frequency Table of # of Contries by Continent
table(dat$Continent)


#basic summary statistics
summary(dat)

#just want a mean
mean(dat$IMR, na.rm=T)
#standard deviation
sd(dat$IMR, na.rm=T)
#Quantiles
quantile(dat$IMR, na.rm=T)

#histogram of the infant mortality rate
hist(dat$IMR, main="Histogram of Infant Mortality Rate")

#Box plot for IMR* Continent
boxplot(IMR~Continent, dat,main="Boxplot of Infant Mortality Rate by continent")

#scatter plot of IMR * TFR
plot(IMR~TFR, data=dat, main="Bivariate Association between TFR and IMR")

#t-test for Africa vs Rest of the world
#Useing the I() funciton, which generates a T/F value, i.e. 2 groups, Africa or Not Africa
t.test(IMR~I(Continent=="Africa"), dat)


#Simple ANOVA model version of a t-test
fit<-lm(IMR~I(Continent=="Africa"), dat)
summary(fit)

#ANOVA model for IMR by continent
fit<-lm(IMR~Continent, dat)
anova(fit)
#Post hoc tests using Tukey comparisons
glht(fit, linfct=mcp(Continent="Tukey"))

#Basic OLS models
#fit the basic regression model
fit1<-lm(IMR ~ TFR + log(GNIPPPperCapitaUSDollars)+ log(PopDensPerSqMile), data=dat)
summary(fit1)
#do some diagnostic plots
plot(fit1)

#Normality of residuals from the first fit?
shapiro.test(rstudent(fit1))
plot(density(rstudent(fit1)))
#test for heteroskedasticity
bptest(fit1)
#variance inflation factors
vif(fit1)

#make White-corrected t-statistics and p values
coeftest(fit1, vcov=vcovHC(fit1, type = "HC0"))

#now we fit a model that includes differences between continents
fit2<-lm(IMR ~ TFR + log(GNIPPPperCapitaUSDollars)+ log(PopDensPerSqMile)+ Continent, data=dat)
summary(fit2)

#We can compare how model 2 performs compared to model 1 by using an F test
anova (fit1, fit2, test="F")


#Now we fit a model with an interaction term between continent and GDP, in R this is easy, all we have to do is use the * operation
fit3<-lm(IMR ~ TFR + log(GNIPPPperCapitaUSDollars)+ log(PopDensPerSqMile), data=dat)
summary(fit3)

#We can compare out model 2 with the interaction model to see if there is a significant difference in how GDP operates across continents
anova (fit2, fit3, test="F")
#This test suggests that there is not a significant interaction between GDP and Continent
```

```{r kable, results='asis'}
#examine a log transformed outcome, this can be easily done within the lm() function
lm2<-lm(log(IMR) ~ TFR + log(GNIPPPperCapitaUSDollars)+ log(PopDensPerSqMile), data=dat)
summary(lm2)

#Normality of errors
shapiro.test(rstudent(lm2))
#Heterosckedasticity test
bptest(lm2)
library(stargazer)
#diagnostic plots
#plot(lm2)
#kable(summary(lm2)$coef, digits = 2, format = "html", caption="Results from model 2", row.names=)

stargazer(fit3, lm2, title="Model 2 results",type="html", align=T, covariate.labels = c("TFR", "lnGDP", "lnDens"))

```


