---
title: "DEM 7283 - Example 4 - Ordinal & Multinomial Logit Models"
author: "coreysparks"
date: "February 9, 2015"
output:
  html_document:
    fig_height: 7
---

This example will cover the use of R functions for fitting Ordinal and Multinomial logit models to complex survey data.

For this example I am using 2011 CDC Behavioral Risk Factor Surveillance System (BRFSS) SMART county data. [Link](http://www.cdc.gov/brfss/smart/smart_2011.htm)

```{r}
#load brfss
library(car)
library(survey)
library(nnet)
load("~/Google Drive/dem7283/data/brfss_11.Rdata")

#The names in the data are very ugly, so I make them less ugly
nams<-names(brfss_11)
head(nams, n=10)
#we see some names are lower case, some are upper and some have a little _ in the first position. This is a nightmare.
newnames<-gsub(pattern = "_",replacement =  "",x =  nams)
names(brfss_11)<-tolower(newnames)

#Our outcome this time is the full spectrum of reported health 
brfss_11$health<-as.factor(ifelse(brfss_11$genhlth%in%c(7,9),NA, brfss_11$genhlth))
brfss_11$health<-relevel(brfss_11$health, ref="1")
brfss_11$health2<-ifelse(brfss_11$genhlth%in%c(7,9),NA, brfss_11$genhlth)

#race/ethnicity
brfss_11$black<-recode(brfss_11$racegr2, recodes="2=1; 9=NA; else=0")
brfss_11$white<-recode(brfss_11$racegr2, recodes="1=1; 9=NA; else=0")
brfss_11$other<-recode(brfss_11$racegr2, recodes="3:4=1; 9=NA; else=0")
brfss_11$hispanic<-recode(brfss_11$racegr2, recodes="5=1; 9=NA; else=0")
brfss_11$race_group<-recode(brfss_11$racegr2, recodes="1='NH white'; 2='NH black'; 3:4='NH other';5='hispanic'; else=NA", as.factor.result = T)
brfss_11$race_group<-relevel(brfss_11$race_group, ref = 'NH white')
#insurance
brfss_11$ins<-ifelse(brfss_11$hlthpln1==1,1,0)

#income grouping
brfss_11$inc<-ifelse(brfss_11$incomg==9, NA, brfss_11$incomg)

#education level
brfss_11$educ<-recode(brfss_11$educa, recodes="1:2='0Prim'; 3='1somehs'; 4='2hsgrad'; 5='3somecol'; 6='4colgrad';9=NA", as.factor.result=T)
#brfss_11$educ<-relevel(brfss_11$educ, ref='0Prim')

#employment
brfss_11$employ<-recode(brfss_11$employ, recodes="1:2='Employed'; 2:6='nilf'; 7='retired'; 8='unable'; else=NA", as.factor.result=T)
brfss_11$employ<-relevel(brfss_11$employ, ref='Employed')

#marital status
brfss_11$marst<-recode(brfss_11$marital, recodes="1='married'; 2='divorced'; 3='widowed'; 4='separated'; 5='nm';6='cohab'; else=NA", as.factor.result=T)
brfss_11$marst<-relevel(brfss_11$marst, ref='married')

#Age cut into intervals
brfss_11$agec<-cut(brfss_11$age, breaks=c(0,24,39,59,79,99))

```

###Analysis
First, we will subset our data to have complete cases for our variables in our model and make our survey design object

```{r}
#First we tell R our survey design
options(survey.lonely.psu = "adjust")
brfss_11<-brfss_11[complete.cases(brfss_11$health,brfss_11$black,brfss_11$educ,brfss_11$agec) ,]
des<-svydesign(ids=~psu, strata=~ststr, weights=~cntywt, data = brfss_11[is.na(brfss_11$cntywt)==F,], nest=T )

```

##Ordinal Regression example
To fit an ordinal logit to survey data in R, we use the `svyolr` fucntion in the survey library. 

```{r}
#Here I fit three nested models for the health outcome
fit.solr1<-svyolr(health~black+hispanic+other,des)
fit.solr2<-svyolr(health~black+hispanic+other+educ,des)
fit.solr3<-svyolr(health~black+hispanic+other+educ+agec,des)

#Show summary of each model and calculate the AIC for each model, even though the AIC is probably not right
summary(fit.solr1);fit.solr1$deviance+2*length(fit.solr1$coefficients)
summary(fit.solr2);fit.solr2$deviance+2*length(fit.solr2$coefficients)
summary(fit.solr3);fit.solr3$deviance+2*length(fit.solr3$coefficients)

#"Examine" proportional odds assumption by fitting logits for each change
ex1<-svyglm(I(health2>1)~black+hispanic+other+educ+agec,des, family="binomial")
ex2<-svyglm(I(health2>2)~black+hispanic+other+educ+agec,des, family="binomial")
ex3<-svyglm(I(health2>3)~black+hispanic+other+educ+agec,des, family="binomial")
ex4<-svyglm(I(health2>4)~black+hispanic+other+educ+agec,des, family="binomial")

#Just a a rough guide, I plot the coefficients of each model. If the proportional odds
#assumption is ok, they all should be "approximately" the same values, they are NOT!!
plot(coef(ex1)[-1], ylim=c(-2, 3), type="l", ylab="Beta", main=c("Comparison of betas for", " proportional odds assumption"))
lines(coef(ex2)[-1], col=2) #not bad
lines(coef(ex3)[-1], col=3) # again, not bad
lines(coef(ex4)[-1], col=4) #not as good, the effect of age are very different for the change between
#fair or better -> poor health


#Just print the odds ratios, 
round(exp(round(rbind(coef(ex1)[-1], coef(ex2)[-1], coef(ex3)[-1], coef(ex4)[-1]),2)),3)

```


###Multinomial Model
Unfortunately, in R there is no multinomial survey design model, so we have to trick it, by creating some replicate weights. This is what Thomas Lumley (the author of the survey package) says to do in this case (and in cases where the type of  model can't be fit otherwise)
**WARNING, THIS LOOKS SCREWY**
```{r}
#create a bootstrap replicate weight set
des2<-as.svrepdesign(des, type="bootstrap" , replicates=10)
#Now I fit the model, this is screwy looking
mfit<-withReplicates(des2, quote(coef(multinom(health~black+hispanic+other+educ+agec,weights=.weights,trace=F ))))
#get the betas
mfitcoef<-data.frame(matrix(attr(attr(mfit, "var"), "means")[-1:-4], nrow=4, ncol=11, byrow=F))
names(mfitcoef)<-names(coef(ex1)[-1])

#odds ratios
round(exp(mfitcoef), 3)

#get the covariance matrix for the betas
vcov<-matrix(attr(mfit, "var"), nrow=48, ncol=48)

#get the z tests by using the coefficients and the standard errors (diag(vcov))
z<-as.vector(mfitcoef/sqrt(diag(vcov)[-1:-4]))
round(z, 2)
#p values
pvals<-round((1-pnorm(abs(as.matrix(z)),0,1))*2,4)
round(pvals, 3)
```
