---
title: "DEM 7283 - Example 5 - Count Data Models"
author: "coreysparks"
date: "February 16, 2015"
output:
  html_document:
    fig_height: 7
---

This example will cover the use of R functions for fitting count data models to complex survey data.

For this example I am using 2011 CDC Behavioral Risk Factor Surveillance System (BRFSS) SMART county data. [Link](http://www.cdc.gov/brfss/smart/smart_2011.htm)

```{r}
#load brfss
library(car)
library(survey)
library(MASS)
library(pscl)
load("~/Google Drive/dem7283/data/brfss_11.Rdata")

#The names in the data are very ugly, so I make them less ugly
nams<-names(brfss_11)
head(nams, n=10)
#we see some names are lower case, some are upper and some have a little _ in the first position. This is a nightmare.
newnames<-gsub(pattern = "_",replacement =  "",x =  nams)
names(brfss_11)<-tolower(newnames)

#Outcome is: Now thinking about your physical health, which includes physical illness and injury, for how many days during the past 30 days was your physical health not good?
brfss_11$healthydays<-ifelse(brfss_11$physhlth %in%c(77,99), NA,ifelse(brfss_11$physhlth==88,0, brfss_11$physhlth))


#race/ethnicity
brfss_11$black<-recode(brfss_11$racegr2, recodes="2=1; 9=NA; else=0")
brfss_11$white<-recode(brfss_11$racegr2, recodes="1=1; 9=NA; else=0")
brfss_11$other<-recode(brfss_11$racegr2, recodes="3:4=1; 9=NA; else=0")
brfss_11$hispanic<-recode(brfss_11$racegr2, recodes="5=1; 9=NA; else=0")
brfss_11$race_group<-recode(brfss_11$racegr2, recodes="1='NH white'; 2='NH black'; 3:4='NH other';5='hispanic'; else=NA", as.factor.result = T)
brfss_11$race_group<-relevel(brfss_11$race_group, ref = 'NH white')
#insurance
brfss_11$ins<-ifelse(brfss_11$hlthpln1==1,1,0)

#income grouping
brfss_11$inc<-ifelse(brfss_11$incomg==9, NA, brfss_11$incomg)

#education level
brfss_11$educ<-recode(brfss_11$educa, recodes="1:2='0Prim'; 3='1somehs'; 4='2hsgrad'; 5='3somecol'; 6='4colgrad';9=NA", as.factor.result=T)
#brfss_11$educ<-relevel(brfss_11$educ, ref='0Prim')

#employment
brfss_11$employ<-recode(brfss_11$employ, recodes="1:2='Employed'; 2:6='nilf'; 7='retired'; 8='unable'; else=NA", as.factor.result=T)
brfss_11$employ<-relevel(brfss_11$employ, ref='Employed')

#marital status
brfss_11$marst<-recode(brfss_11$marital, recodes="1='married'; 2='divorced'; 3='widowed'; 4='separated'; 5='nm';6='cohab'; else=NA", as.factor.result=T)
brfss_11$marst<-relevel(brfss_11$marst, ref='married')

#Age cut into intervals
brfss_11$agec<-cut(brfss_11$age, breaks=c(0,24,39,59,79,99))

```

###Analysis
First, we will subset our data to have complete cases for our variables in our model and make our survey design object

```{r}
#Here I keep complete cases on my key variables, AND only cases for TX, just for speed (the suvey procedures can run for a long time)
brfss_11<-brfss_11[is.na(brfss_11$healthydays)==F&is.na(brfss_11$cntywt)==F&is.na(brfss_11$black)==F&is.na(brfss_11$educ)==F&brfss_11$state==48,]

#FOR THE BRFSS, R GAVE ME A WARNING AND I NEEDED TO ADD:
options(survey.lonely.psu = "adjust")
#YOU MAY NOT NEED TO DO THIS!!!!
#First we tell R our survey design
options(survey.lonely.psu = "adjust")
des<-svydesign(ids=~psu, strata=~ststr, weights=~cntywt, data = brfss_11[is.na(brfss_11$cntywt)==F,], nest=T )

#Make a replicate weight design, to fit models not in the survey package
repdes<-as.svrepdesign(des, type="bootstrap", replicates=100)
```

##Poisson regression example
To fit a Poisson GLM to survey data in R, we use the `svyglm` fucntion in the survey library. 

```{r}
#First I do some simple descriptives
svyhist(~healthydays, des)
svyby(~healthydays, ~race_group+educ, des, svymean)
svyby(~healthydays, ~race_group+educ, repdes, svymean)
#Poisson glm fit to survey data
fit1<-svyglm(healthydays~race_group+educ+agec, design=des, family=poisson)
summary(fit1)
#here are the poisson model "risk ratios", which just show the change in the mean
round(exp(summary(fit1)$coef[-1,1]), 3)
```

Now, R will not fit the other count data models described in the notes, so, we will fit them using sample weights only, then calculate the robust standard errors. We standardize the weights to equal the sample size, as opposed to the population size by dividing each person's weight by the mean weight.

```{r}
#First, I define a function to get the clustered, or robust standard errors. This function effectively controls for the within-strata homogeneity when calculateing the se's for the betas. 

#I stole this from: http://drewdimmery.com/robust-ses-in-r/
#and http://people.su.se/~ma/clustering.pdf
#I also added a correction to use this with the hurdle and zero-inflated models
#This is how stata gets robust se's

clx2 <-   function(fm, dfcw,  cluster){
    # R-codes (www.r-project.org) for computing
    # clustered-standard errors. Mahmood Arai, Jan 26, 2008.
    
    # The arguments of the function are:
    # fitted model, cluster1
    # You need to install libraries `sandwich' and `lmtest'
    
    # reweighting the var-cov matrix for the within model
    require(sandwich);require(lmtest)
    if(class(fm)=="zeroinfl"|class(fm)=="hurdle") {
    M <- length(unique(cluster))   
    N <- length(cluster)           
    K <- dim(fm$vcov)[1]        #here is the rank from the zero inflated fits             
    dfc <- (M/(M-1))*((N-1)/(N-K))  
    uj  <- apply(estfun(fm),2, function(x) tapply(x, cluster, sum));
    vcovCL <- dfc[1]*sandwich(fm, meat=crossprod(uj)/N)*dfcw #fix a length problem in dfc
    list(summary=coeftest(fm, vcovCL))}
    else if(class(fm)!="zeroinfl"){
    M <- length(unique(cluster))
    N <- length(cluster)
    K <- fm$rank
    dfc <- (M/(M - 1)) * ((N - 1)/(N - K))
    uj <- apply(estfun(fm), 2, function(x) tapply(x, cluster, sum));
    rcse.cov <- dfc * sandwich(fm, meat = crossprod(uj)/N)
    rcse.se <- coeftest(fm, rcse.cov)
    return(list( rcse.se))}
}


#Fit poisson, and compare it to the fit from the survey design
#Fit the Poisson GLM
fit.pois<-glm(healthydays~race_group+educ+agec, data=brfss_11, weights=cntywt/mean(cntywt), family=poisson)
fit.pois2<-glm(healthydays~race_group+educ, data=brfss_11, weights=cntywt/mean(cntywt), family=poisson)
anova(fit.pois2, fit.pois)
summary(fit.pois)
#here are the poisson model "risk ratios"
round(exp(summary(fit.pois)$coef[-1,1]), 3)
#I use psu as the clustering variable, this is legal.
#Here is the Poisson model
clx2(fit.pois, 1, brfss_11$psu)
#which looks nearly identical to the output using the survey design from above

#Fit the Negative Binomial GLM
fit.nb2<-withReplicates(design = repdes, quote(coef(glm.nb(healthydays~race_group+educ+agec,  weights=.weights))))
#You can use the replicate design for the NB model, as we did with the multinomial model last week:
fit.nb<-glm.nb(healthydays~race_group+educ+agec, data=brfss_11, weights=cntywt/mean(cntywt))
fit.nb
summary(fit.nb)
clx2(fit.nb, 1, brfss_11$psu)

#Fit the Zero Inflated Poisson Mixture Model
fit.z1<-withReplicates(design = repdes, quote(coef(zeroinfl(healthydays~race_group+educ+agec|1,dist="poisson",  weights=.weights))))

fit.zip<-zeroinfl(healthydays~race_group+educ+agec|1, data=brfss_11, weights=cntywt/mean(cntywt), dist="poisson")
fit.z1
summary(fit.zip)
clx2(fit.zip, 1, brfss_11$psu)


#Fit the Zero Inflated Negative Binomial Mixture Model
fit.zinb<-zeroinfl(healthydays~race_group+educ+agec|agec, data=brfss_11, weights=cntywt/mean(cntywt), dist="negbin")
summary(fit.zinb)
clx2(fit.zinb, 1, brfss_11$psu)

#Fit the Poisson Hurdle Model
fit.hp<-hurdle(healthydays~race_group+educ+agec|agec, data=brfss_11, weights=cntywt/mean(cntywt), dist="poisson")
summary(fit.hp)
clx2(fit.hp, 1, brfss_11$psu)

#Fit the Negative Binomial Hurdle Model
fit.hnb<-hurdle(healthydays~race_group+educ+agec|agec, data=brfss_11, weights=cntywt/mean(cntywt), dist="negbin")
summary(fit.hnb)
clx2(fit.hnb, 1, brfss_11$psu)


#Collect all the model AICs
AICS<-c(AIC(fit.pois),
AIC(fit.nb),
AIC(fit.zip),
AIC(fit.zinb),
AIC(fit.hp),
AIC(fit.hnb))

#Plot them here are are looking for the smallest value
plot(AICS, type="p",pch=24,bg="grey", cex=2, ylab="AIC",axes=T,xaxt="n", xlab="", ylim=c(min(AICS)-5000, max(AICS)+5000))
title(main="AIC Values for Alternative Models")
axis(1, at=1:6,labels=F) #6= number of models
labels<-c("Poisson model", "NB Model","ZI-Pois", "ZI-NebBin","HurdlePois", "HurdleNB" )
text(1:6, par("usr")[3]-.25, srt=45, adj=1, labels=labels, xpd=T)
mtext(side=1, text="Model Specification", line=3)
symbols(x= which.min(AICS), y=AICS[which.min(AICS)], circles=5, fg=2,lwd=3,add=T)

AICS[which.min(AICS)]
#looks like the negative binomimal fits best  


#get a series of predicted counts of days for different "types" of people for each model
dat<-expand.grid(race_group=levels(brfss_11$race_group), educ=levels(brfss_11$educ), agec=levels(brfss_11$agec))

#generate the fitted values
fitted<-predict(fit.zinb, newdat=dat,type="response")
#add the values to the fake data
dat$fitted.days<-round(fitted, 3)

#Print the fitted number of days
head(dat, n=20)


```
