---
title: "Week 4: Basic Hierarchical Models - Other Topics"
output: html_document
---

Here, we will talk about a few more miscellaneous topics: GLMM's and Cross level interactions
The data I will use here is from the Behavioral Risk Factor Surveillance System (BRFSS) SMART county data
First we load our data
```{r}
library(lme4)
library(arm)
library(car)
library(MASS)
library(ordinal)

#load brfss
load("~/Google Drive/dem7903_App_Hier/data/brfss_11.Rdata")
nams<-names(brfss_11)
newnames<-gsub("_", "", nams)
names(brfss_11)<-tolower(newnames)

brfss_11$statefip<-sprintf("%02d", brfss_11$state )
brfss_11$cofip<-sprintf("%03d", brfss_11$cnty )
brfss_11$cofips<-paste(brfss_11$statefip, brfss_11$cofip, sep="")

#Outcome is: Now thinking about your physical health, which includes physical illness and injury, for how many days during the past 30 days was your physical health not good?
brfss_11$healthydays<-ifelse(brfss_11$physhlth %in%c(77,99), NA,ifelse(brfss_11$physhlth==88,0, brfss_11$physhlth))

#just for brevity, I just select TX respondents with non missing weights
brfss_11<-brfss_11[is.na(brfss_11$cntywt)==F,]
#insurance
brfss_11$ins<-ifelse(brfss_11$hlthpln1==1,1,0)
#smoking currently
brfss_11$smoke<-recode(brfss_11$smoker3, recodes="1:2=1; 3:4=0; else=NA")
#low physical activity
brfss_11$lowact<-recode(brfss_11$pacat, recodes="1:2=0; 3:4=1; else=NA")
#high blood pressure
brfss_11$hbp<-recode(brfss_11$bphigh4, recodes="1=1; 2:4=0; else=NA")
#high cholesterol
brfss_11$hc<-recode(brfss_11$toldhi2, recodes="1=1; 2=0; else=NA")
#bmi
brfss_11$bmi<-ifelse(is.na(brfss_11$bmi5)==T, NA, brfss_11$bmi5/100)
#poor or fair health
brfss_11$badhealth<-ifelse(brfss_11$genhlth %in% c(4,5),1,0)

#race
brfss_11$black<-recode(brfss_11$racegr2, recodes="2=1; 9=NA; else=0", as.factor.result=T)
brfss_11$white<-recode(brfss_11$racegr2, recodes="1=1; 9=NA; else=0", as.factor.result=T)
brfss_11$other<-recode(brfss_11$racegr2, recodes="3:4=1; 9=NA; else=0", as.factor.result=T)
brfss_11$hispanic<-recode(brfss_11$racegr2, recodes="5=1; 9=NA; else=0", as.factor.result=T)

#have a personal doctor?
brfss_11$doc<-recode(brfss_11$persdoc2, recodes="1:2=1; 3=0; else=NA", as.factor.result=F)

#needed care in last year but couldn't get it because of cost
brfss_11$medacc<-recode(brfss_11$medcost, recodes="1=1;2=0;else=NA")

#education level
brfss_11$lths<-recode(brfss_11$educa, recodes="1:3=1;9=NA; else=0", as.factor.result=F)
brfss_11$coll<-recode(brfss_11$educa, recodes="5:6=1;9=NA; else=0", as.factor.result=F)

#employment
brfss_11$employ<-recode(brfss_11$employ, recodes="1:2='Employed'; 2:6='nilf'; 7='retired'; 8='unable'; else=NA", as.factor.result=T)
brfss_11$employ<-relevel(brfss_11$employ, ref='Employed')

#marital status
brfss_11$marst<-recode(brfss_11$marital, recodes="1='married'; 2='divorced'; 3='widowed'; 4='separated'; 5='nm';6='cohab'; else=NA", as.factor.result=T)
brfss_11$marst<-relevel(brfss_11$marst, ref='married')

#income
brfss_11$inc<-as.factor(ifelse(brfss_11$incomg==9, NA, brfss_11$incomg))

#Age cut into intervals
brfss_11$agec<-cut(brfss_11$age, breaks=c(0,24,39,59,79,99))

```

Now we will begin fitting the multilevel regression model with the county
that the person lives in being the higher level

I will also add in some Census variables from the ACS 2010 5 Year estimates
load in ACS data from Factfinder
```{r}
acsecon<-read.csv("~/Google Drive/dem7903_App_Hier/data/aff_download/ACS_10_5YR_DP03_with_ann.csv")
acsecon$povrate<-acsecon[, "HC03_VC156"]
acsecon$unemployed<-acsecon[, "HC03_VC13"]
acsecon$cofips<-substr(acsecon$GEO.id, 10,14)
acsecon<-acsecon[, c("cofips", "povrate", "unemployed")]

joindata<-acsecon
head(acsecon)
```


I also have segregation measures at the count level that I made, I will merge these in too
As a rule, when including continuous predictors at the higher level, you should always standardize them
This is done for 2 reasons, first is that it makes the computer happy, second is that it makes the interpretation of the higher level variable easier.


```{r}
load("~/Google Drive/dem7903_App_Hier/data/segregation_county.Rdata")
joindata<-merge(joindata, county_dat, by.x="cofips", by.y="cofips", all.x=T)

joindata$povz<-scale(joindata$povrate, center=T, scale=T)
joindata$unempz<-scale(joindata$unemployed, center=T, scale=T)
joindata$diss.z<-scale(joindata$dissim_wb, center=T, scale=T)
joindata$iso.z<-scale(joindata$isolation_b, center=T, scale=T)
joindata$theil.z<-scale(joindata$TheileH, center=T, scale=T)

head(joindata)
#and merge the data back to the kids data
merged<-merge(x=brfss_11, y=joindata, by.x="cofips", by.y="cofips", all.x=T)
merged$bmiz<-scale(merged$bmi, center=T, scale=T)
merged$agez<-scale(merged$age, center=T, scale=T)

```


Here I fit a binomial model (binary logistic regression) for healthcare access (1= has a regular primary care physician, 0=does not)
First a simple random intercept model, then a model with a higher level variable, then a cross level interaction model

```{r}
#################
#### GLMM's #####
#################
#I subset the data to only have states with more than 8 counties sampled
tapply(merged$cofips, merged$state, function(x) length(unique(x))>5)
merged<-merged[merged$statefip%in%c("06","08","23","30", "31", "25","39","45" ,"48","49"),]

fit.mix.bin<-glmer(doc~agez+lths+coll+black+hispanic+other+(1|cofips), family=binomial, data=merged)
summary(fit.mix.bin)

#Then I do a multi-level model where I include the poverty rate and the segregation index at the county level
fit.mix.bin.ml<-glmer(doc~agez+lths+coll+black+hispanic+other+povz+(1|cofips), family=binomial, data=merged)
summary(fit.mix.bin.ml)
anova(fit.mix.bin, fit.mix.bin.ml)
#Little evidence of poverty effect

#Cross level interaction model between poverty and race and segregation and race
#You simply include a term of individualvar*highervar
fit.mix.bin.cl<-glmer(doc~agez+lths+coll+(black+hispanic+other)*povz+(1|cofips), family=binomial, data=merged)
summary(fit.mix.bin.cl)
anova(fit.mix.bin.ml, fit.mix.bin.cl)
#Cross level interaction is not significant
```

Here is an example of using a Poisson outcome, number of healthy days reported in the last month
```{r}
fit.pois<-glmer(healthydays~(black+hispanic+other)+lths+coll+agez+badhealth+(1|cofips),family=poisson, data=merged)
summary(fit.pois)
```

Perhaps a negative binomial model would be more appropriate, as our outcome is very overdispersed:
```{r}
fit.nb<-glmer.nb(healthydays~(black+hispanic+other)+lths+coll+agez+badhealth+(1|cofips), data=merged)
summary(fit.nb)

```


GLMMs can also be used on ordinal outcomes, for example an ordinal logit:
```{r}
#Ordinal Logit GLMM
merged$hlth<-recode(merged$genhlth, recodes="1=1; 2=2; 3=3; 4=4;5=5; else=NA", as.factor.result=T)
table(merged$hlth)

fit.mix.clmm<-clmm(hlth~agez+lths+coll+black+hispanic+other+theil.z+povz+(1|cofips), data=merged)
summary(fit.mix.clmm)

```


how do I get ICC from one of these? For the binomial model, the residual variance is a constant of $\pi^{2}/3$
For E.g. from the first model:
```{r}
.142/(.142+ (pi^2)/3)

```
