---
title: 'Week 3: Basic Hierarchical Models - Random Intercepts and Slopes'
output:
  html_document:
    highlight: tango
---

First we load our data
```{r}
load("C:/Users/ozd504//Google Drive/dem7903_App_Hier/data/eclsk.Rdata")
names(eclsk)<-tolower(names(eclsk))
library (car)
library(lme4)
library(arm)
#get out only the variables I'm going to use for this example

myvars<-c("gender", "kurban_r", "race", "w1povrty", "wkmomed","p2homecm", "p2cover", "p2curmar", "p2sprhhm", "c2r4mtsc", "p2dentis", "s2_id", "s2kpupri","s2kminor" )
#subset the data
eclsk.sub<-eclsk[,myvars]
rm(eclsk); gc()
```


Next, I do some recoding of variables using a mixture of the ` ifelse()` function and the `recode ()` function.
```{r}
#recode our outcomes, the  first is the child's math standardized test score  in Kindergarten
eclsk.sub$math<-ifelse(eclsk.sub$c2r4mtsc<0, NA, eclsk.sub$c2r4mtsc)

#the second outcome is whether each child has seen a dentist within the last year
eclsk.sub$dentist<-recode(eclsk.sub$p2dentis, recodes = "2:3=1; -1:-9= NA; else=0")

#First we recode some Child characteristics
#Child's sex: recode as male =1
eclsk.sub$male<-recode(eclsk.sub$gender, recodes="1=1; 2=0; -9=NA")

#Recode race with white, non Hispanic as reference using dummy vars
eclsk.sub$hisp<-recode (eclsk.sub$race, recodes="3:4=1;-9=NA; else=0")
eclsk.sub$black<-recode (eclsk.sub$race, recodes="2=1;-9=NA; else=0")
eclsk.sub$asian<-recode (eclsk.sub$race, recodes="5=1;-9=NA; else=0")
eclsk.sub$nahn<-recode (eclsk.sub$race, recodes="6:7=1;-9=NA; else=0")
eclsk.sub$other<-recode (eclsk.sub$race, recodes="8=1;-9=NA; else=0")

#insurance coverage
eclsk.sub$covered<-recode(eclsk.sub$p2cover, recodes="1=1; -1:-9=NA; else=0")

#Then we recode some parent/mother characteristics
#Mother's education, recode as 2 dummys with HS = reference
eclsk.sub$mlths<-recode(eclsk.sub$wkmomed, recodes = "1:2=1; 3:9=0; else = NA")
eclsk.sub$mgths<-recode(eclsk.sub$wkmomed, recodes = "1:3=0; 4:9=1; else =NA") 

#marital status, recode as 2 dummys, ref= married
eclsk.sub$single<-recode(eclsk.sub$p2curmar, recodes="5=1; -7:-9=NA; else=0")
eclsk.sub$notmar<-recode(eclsk.sub$p2curmar, recodes="2:4=1; -7:-9=NA; else=0")


#Then we do some household level variables
#Urban residence = 1
eclsk.sub$urban<-recode(eclsk.sub$kurban_r, recodes = "1:2=1; 3=0")

#poverty level in poverty = 1
eclsk.sub$pov<-recode(eclsk.sub$w1povrty , recodes ="1=1; 2=0")

#Household size
eclsk.sub$hhsize<-eclsk.sub$p2sprhhm

#school is private
eclsk.sub$privsch<-ifelse(eclsk.sub$s2kpupri==1, 1, 0)

#school has >50% minority student body
eclsk.sub$minorsch<-recode(eclsk.sub$s2kminor, recodes="1:3=0; 4:5=1; else=NA" )
```

```{r}
#Show the first few lines of the data
head(eclsk.sub)
```


The First model will fit a model that considers the individual and family level variables 
and a random intercept only
```{r}
fit1<-lmer(math~male+hisp+black+asian+nahn+other+mlths+mgths+single+notmar+urban+pov+hhsize+(1|s2_id), data=eclsk.sub, REML=T)

display(fit1)
```


The second model considers both random intercepts and random slopes, in this case, i'm only considering a random slope for poverty status

```{r}
fit2<-lmer(math~male+hisp+black+asian+nahn+other+mlths+mgths+single+notmar+urban+pov+hhsize+(1+pov|s2_id), data = eclsk.sub, REML = T)
display(fit2)

anova (fit1, fit2)
```
In this case, fitting the random slope for poverty isn't adding anything to the model
I interpret this as kids living in housholds below the poverty line face a systematic disadvantage, regardless of school

Now, let's plot the regression lines for each school!
```{r}
#get the random effects
rancoefs2<-ranef(fit2)

#here are the first 10 schools random effects
head(rancoefs2$s2_id, n=10)

#Histograms 
par(mfrow=c(1,2))
hist(rancoefs2$s2_id[,"(Intercept)"], main = "Distribution of Random Intercepts")
hist(rancoefs2$s2_id[,"pov"], main = "Distribution of Random Slopes")


#Plot of random slopes, I iterate over all schools, drawing lines
par(mfrow=c(1,1))
fixef(fit2)
summary(fixef(fit2)[1]+rancoefs2$s2_id)
plot(NULL, ylim=c(35, 65), xlim=c(0,1),ylab="Math Score", xlab="HH Poverty Status")
title (main="Random Slope and Intercept Model")
cols=sample(rainbow(n=50), size = dim(rancoefs2$s2_id)[1], replace = T)
for (i in 1: dim(rancoefs2$s2_id)[1]){
  
  abline(a=fixef(fit2)[1]+rancoefs2$s2_id[[1]][i], b=fixef(fit2)[13]+rancoefs2$s2_id[[2]][i], col=cols[i],lwd=.5 )
}

```

Here is the ICC from fit 2
```{r}
VarCorr(fit2)
VarCorr(fit2)$s2_id[4]/((attr(VarCorr(fit1), "sc")^2)+VarCorr(fit2)$s2_id[1]+VarCorr(fit2)$s2_id[4])
#the "sc" attribute is the residual variance
```

