---
title: "Week 1: Linear Model Review"
output: html_document
---

First, we will show an example of fitting the basic linear regression model and the basic logistic regression model. 

First we load our data, if you have an Rdata file, use load. If your data are in a different format, you can use the ` foreign ` library. R can read SPSS, Stata, SAS transport, Excel, CSV, DBF and other formats.

```{r}
load("~/Google Drive/dem7903_App_Hier/data/eclsk.Rdata")
#we see what its called by listing the objects in memory
ls()
```
We see our data are called eclsk

R is case sensitive, and the names of the data are ugly, so I convert them to all lower case:
```{r}
names(eclsk)<-tolower(names(eclsk))
```


Next, we will load the car and lmtest libraries, you may need to install them by : 
`install.packages(c("car", "lmtest"))`

```{r}
library (car)
library(lmtest)
```

The ECLS-K has around 18,000 variables, and we don't need all of these. 
Next make a subset of the data that only contains what we need, I know these variable names already. 
```{r}
myvars<-c("gender", "kurban_r", "race", "w1povrty", "wkmomed","p2homecm", "p2cover", "p2curmar", "p2sprhhm", "c2r4mtsc", "p2dentis")
```

In R, everthing is a matrix, for example: `x[rows,columns]`
so this code: 
```{r}
eclsk.sub<-eclsk[,myvars]
```
Will select every column that matches one of the names I defined in myvars above

Next, I do some recoding of variables using a mixture of the ` ifelse()` function and the `recode ()` function.
```{r}
#recode our outcomes, the  first is the child's math standardized test score  in Kindergarten
eclsk.sub$math<-ifelse(eclsk.sub$c2r4mtsc<0, NA, eclsk.sub$c2r4mtsc)

#the second outcome is whether each child has seen a dentist within the last year
eclsk.sub$dentist<-recode(eclsk$p2dentis, recodes = "2:3=1; -1:-9= NA; else=0")

#First we recode some Child characteristics
#Child's sex: recode as male =1
eclsk.sub$male<-recode(eclsk.sub$gender, recodes="1=1; 2=0; -9=NA")

#Recode race with white, non Hispanic as reference using dummy vars
eclsk.sub$hisp<-recode (eclsk.sub$race, recodes="3:4=1;-9=NA; else=0")
eclsk.sub$black<-recode (eclsk.sub$race, recodes="2=1;-9=NA; else=0")
eclsk.sub$asian<-recode (eclsk.sub$race, recodes="5=1;-9=NA; else=0")
eclsk.sub$nahn<-recode (eclsk.sub$race, recodes="6:7=1;-9=NA; else=0")
eclsk.sub$other<-recode (eclsk.sub$race, recodes="8=1;-9=NA; else=0")

#insurance coverage
eclsk.sub$covered<-recode(eclsk.sub$p2cover, recodes="1=1; -1:-9=NA; else=0")

#Then we recode some parent/mother characteristics
#Mother's education, recode as 2 dummys with HS = reference
eclsk.sub$mlths<-recode(eclsk.sub$wkmomed, recodes = "1:2=1; 3:9=0; else = NA")
eclsk.sub$mgths<-recode(eclsk.sub$wkmomed, recodes = "1:3=0; 4:9=1; else =NA") 

#marital status, recode as 2 dummys, ref= married
eclsk.sub$single<-recode(eclsk.sub$p2curmar, recodes="5=1; -7:-9=NA; else=0")
eclsk.sub$notmar<-recode(eclsk.sub$p2curmar, recodes="2:4=1; -7:-9=NA; else=0")


#Then we do some household level variables
#Urban residence = 1
eclsk.sub$urban<-recode(eclsk.sub$kurban_r, recodes = "1:2=1; 3=0")

#poverty level in poverty = 1
eclsk.sub$pov<-recode(eclsk.sub$w1povrty , recodes ="1=1; 2=0")

#Household size
eclsk.sub$hhsize<-eclsk.sub$p2sprhhm
```

Then I subset the data file to only have complete cases for our variables. I do this because I want to do model comparison below, and models need to be fit to the same data.
```{r}
eclsk.sub<-subset(eclsk.sub, complete.cases(eclsk.sub))
dim(eclsk.sub)
#Show the first few lines of the data
head(eclsk.sub)
```


## Our first example uses an OLS Regression model for a continous outcome
Now we fit our linear regression model using the `lm() ` function, for linear model.
we fit 3 models 
1. Model 1 = child characteristics
2. Model 2 = Model 1 + Parent characteristics
3. Model 3 = Model 2 + household characteristics

```{r}
fit1<-lm(math~male+hisp+black+asian+nahn+other, data=eclsk.sub)

fit2<-lm(math~male+hisp+black+asian+nahn+other+mlths+mgths+single+notmar, data=eclsk.sub)

fit3<-lm(math~male+hisp+black+asian+nahn+other+mlths+mgths+single+notmar+urban+pov+hhsize, data=eclsk.sub)

```

Next, we examine our results for each model. This is how we get our parameter estimates and p-values etc.
```{r}
summary(fit1)
summary(fit2)
summary(fit3)
```

Then we compare our models  using F tests to compare the changes in model fit:
```{r}
anova (fit1, fit2, test="F")
anova (fit2, fit3, test="F")
```

Then, we like to examine the relative changes in model fit using the AIC for models:
```{r}
AIC (fit1)
AIC (fit2)
AIC (fit3)
```


## Next we examine a logistic regression model for our dichotomous outcome

We use the `glm()` function with the `family=binomial` option to specify a logistic regression model.
```{r}

fit1.lr<-glm(dentist~male+hisp+black+asian+nahn+other+covered, data=eclsk.sub, family=binomial)

fit2.lr<-glm(dentist~male+hisp+black+asian+nahn+other+covered+mlths+mgths+single+notmar, data=eclsk.sub, family=binomial)

fit3.lr<-glm(dentist~male+hisp+black+asian+nahn+other+covered+mlths+mgths+single+notmar+urban+pov+hhsize, data=eclsk.sub, family=binomial)
```

Then we display the model results
```{r }
summary(fit1.lr)
summary(fit2.lr)
summary(fit3.lr)
```

We can do plots of the odds ratios very easily using the `sjp.glm` function in the `sjPlot` library
Here I just do it for the last model.
```{r, fig.width=7, fig.height=9}
library(sjPlot)
sjp.glm(fit3.lr, sortOdds = F)
```