---
title: "Week 2: Basic Hierarchical Models - Random Intercepts"
output: html_document
---

First we load our data
```{r}
load("~/Google Drive/dem7903_App_Hier/data/eclsk.Rdata")
names(eclsk)<-tolower(names(eclsk))
library (car)
library(lmtest)

#get out only the variables I'm going to use for this example

myvars<-c("gender", "kurban_r", "race", "w1povrty", "wkmomed","p2homecm", "p2cover", "p2curmar", "p2sprhhm", "c2r4mtsc", "p2dentis", "s2_id", "s2kpupri","s2kminor" )
#subset the data
eclsk.sub<-eclsk[,myvars]
rm(eclsk); gc()
```


Next, I do some recoding of variables using a mixture of the ` ifelse()` function and the `recode ()` function.
```{r}
#recode our outcomes, the  first is the child's math standardized test score  in Kindergarten
eclsk.sub$math<-ifelse(eclsk.sub$c2r4mtsc<0, NA, eclsk.sub$c2r4mtsc)

#the second outcome is whether each child has seen a dentist within the last year
eclsk.sub$dentist<-recode(eclsk.sub$p2dentis, recodes = "2:3=1; -1:-9= NA; else=0")

#First we recode some Child characteristics
#Child's sex: recode as male =1
eclsk.sub$male<-recode(eclsk.sub$gender, recodes="1=1; 2=0; -9=NA")

#Recode race with white, non Hispanic as reference using dummy vars
eclsk.sub$hisp<-recode (eclsk.sub$race, recodes="3:4=1;-9=NA; else=0")
eclsk.sub$black<-recode (eclsk.sub$race, recodes="2=1;-9=NA; else=0")
eclsk.sub$asian<-recode (eclsk.sub$race, recodes="5=1;-9=NA; else=0")
eclsk.sub$nahn<-recode (eclsk.sub$race, recodes="6:7=1;-9=NA; else=0")
eclsk.sub$other<-recode (eclsk.sub$race, recodes="8=1;-9=NA; else=0")

#insurance coverage
eclsk.sub$covered<-recode(eclsk.sub$p2cover, recodes="1=1; -1:-9=NA; else=0")

#Then we recode some parent/mother characteristics
#Mother's education, recode as 2 dummys with HS = reference
eclsk.sub$mlths<-recode(eclsk.sub$wkmomed, recodes = "1:2=1; 3:9=0; else = NA")
eclsk.sub$mgths<-recode(eclsk.sub$wkmomed, recodes = "1:3=0; 4:9=1; else =NA") 

#marital status, recode as 2 dummys, ref= married
eclsk.sub$single<-recode(eclsk.sub$p2curmar, recodes="5=1; -7:-9=NA; else=0")
eclsk.sub$notmar<-recode(eclsk.sub$p2curmar, recodes="2:4=1; -7:-9=NA; else=0")


#Then we do some household level variables
#Urban residence = 1
eclsk.sub$urban<-recode(eclsk.sub$kurban_r, recodes = "1:2=1; 3=0")

#poverty level in poverty = 1
eclsk.sub$pov<-recode(eclsk.sub$w1povrty , recodes ="1=1; 2=0")

#Household size
eclsk.sub$hhsize<-eclsk.sub$p2sprhhm

#school is private
eclsk.sub$privsch<-ifelse(eclsk.sub$s2kpupri==1, 1, 0)

#school has >50% minority student body
eclsk.sub$minorsch<-recode(eclsk.sub$s2kminor, recodes="1:3=0; 4:5=1; else=NA" )
```

```{r}
#Show the first few lines of the data
head(eclsk.sub)
```

First, I want to test for variation in my outcome across the  schools,
this is ALWAYS THE FIRST STEP!!!!!  If there is no variation across your groups, why do
a hierarchical model??

```{r}
fit0<-lm(math~factor(s2_id), data=eclsk.sub)
anova(fit0)
```
#Yes!

My logic of fitting a hierarchical model here is that I  have `r length(unique(eclsk.sub$s2_id))` schools in the data. How would I possibly choose one of them as a reference group?  Indeed, I don't have a complete sampling of schools, so a fixed effects model is not really appropriate.

Next, I proceed to fit the random intercept model and examine it
There are lots of ways to fit hierarhical models in R, but I like the `lme4` library because the methods it implements are the more robust type of methods for fitting these models *more on this later*

```{r}
library(lme4) #You will need to install this if you haven't done so already
```

This first model will only fit the means of the schools, I'm doing this to get a "null" model fit, so I have a basis to compare  my other models to.
```{r}
fit1<-lmer(math~1+(1|s2_id), data = eclsk.sub, REML = T)
summary(fit1)
```

The next model will fit a model that considers the individual and family level variables
```{r}
fit2<-lmer(math~male+hisp+black+asian+nahn+other+mlths+mgths+single+notmar+urban+pov+hhsize+(1|s2_id), data=eclsk.sub, REML=T)

summary(fit2)
```

We then compare these two fits using a Likelihood ratio test
```{r}
anova(fit1, fit2)
```

The second model is a better fit than the first, so it's not just the school in and of itself that matters


Finally, I extract the ICCs from each fit, to do this, we use the cryptic `VarCorr()` function that works with fits from `lmer()`

Here is the ICC from fit 1
```{r}
VarCorr(fit1)$s2_id[1]/((attr(VarCorr(fit1), "sc")^2)+VarCorr(fit1)$s2_id[1])
#the "sc" attribute is the residual variance, who knew?
```

And model 2

```{r}
VarCorr(fit2)$s2_id[1]/((attr(VarCorr(fit2), "sc")^2)+VarCorr(fit2)$s2_id[1])
```

Basically what this shows us is that we have reduced the variability between the schools by controlling for children's characteristics

